Making Sequential Programs Deterministic

### 2019-04-04 design.tex
Edits through section 2.5

@@

\begin{figure}
    $\infer[\rulename{ordernondet polyup}]
    {\up{\tau} = \<NonDet> \ \beta}
    {\tau = \<OrderNonDet> \ \beta
        & \wellformed{\tau}
    }$

    $\infer[\rulename{polyup}]
    {\up{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\<Det>, \<NonDet>\}
        & \wellformed{\tau}
    }$

    $\infer[\rulename{ordernondet polydown}]
    {\down{\tau} = \<Det> \ \beta}
    {\tau = \<OrderNonDet> \ \beta
        & \wellformed{\tau}
    }$

    $\infer[\rulename{polydown}]
    {\down{\tau} = \kappa \ \beta}
    {\tau =\kappa \ \beta
        & \kappa \in \{\<Det>, \<NonDet>\}
        & \wellformed{\tau}
    }$
\caption{Polymorphic resolution of $\uparrow$ and $\downarrow$ operators ($\uparrow$ is overloaded for types and qualifiers)}
\todo{I think this needs to do something to nested collections.  Or maybe
  just clarify that if there is a nested collection such that this would
  yield an invalid type, then the instantiation does not apply.  That might
  be cleanest.}
\label{fig:poly-resolutions}
\end{figure}

@@

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
\TheDeterminismChecker performs local type inference within method bodies.
It does not perform whole-program type inference.
This makes separate compilation possible.
It forces programmers to write specifications (type qualifiers) on methods,
which is good style and valuable documentation.

@@

In addition to assignments, some method calls refine the types of their
arguments.  One example is the \<sort> routine, as shown in 
as shown in \cref{fig-sorting}.  One notable fact about the \rulename{sort}
rule is that it requires the type argument to be deterministic (type
qualifier \<Det>).  This is necessary because of two constraints.
First, refinement must not change the type argument, because of invariant
type argument subtyping.
Second, the only possible type argument for a \<Det> collection is \<Det>.

It is not permitted to

@@

  $\infer[\rulename{sort}]
  {\Gamma \vdash x : \<Det List>\angles{\<Det>\ \beta}}
  {\Gamma \vdash x : \<OrderNonDet List>\angles{\<Det>\ \beta} & \<sort(x)>}$

@@

    \caption{Sorting type refinement rules.  \<sort> is a method that works
    by side effect and changes the value---and the type---of its receiver.}
\todo{These don't capture the notion of ``the type \emph{after} the sort call''.}
    \label{fig-sorting}
\end{figure}
\TheDeterminismChecker type refines a receiver of type \codeid{OrderNonDet List} to a \codeid{Det List} if all the
type argument of this \codeid{List} have the type qualifier \codeid{Det} (Rule \rulename{sort} in \cref{fig-sorting}).
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{OrderNonDet List<Det Set<Det String>>}, it will
be type refined to \codeid{Det List<Det Set<Det String>>}. However a receiver of type \codeid{OrderNonDet List<OrderNonDet Set<Det String>>} will not be type refined.

>>>>

\begin{figure}
    $\infer{\up{\<NonDet>} = \<NonDet>}{}
    \quad
    \infer{\up{\<OrderNonDet>} = \<NonDet>}{}
    \quad
    \infer[\rulename{polyup}]{\up{\<Det>} = \<Det>}{}
    $

    $\infer{\down{\<NonDet>} = \<NonDet>}{}
    \quad
    \infer{\down{\<OrderNonDet>} = \<Det>}{}
    \quad
    \infer[\rulename{polydown}]{\down{\<Det>} = \<Det>}{}
    $

\caption{Polymorphic resolution of $\uparrow$ and $\downarrow$ operators.  $\uparrow$ is overloaded for types and qualifiers.}
\label{fig:poly-resolutions}
\end{figure}

@@

@@

\todo{TODO: Need to work on this section.}

In the annotated JDK, the type of get is:

\begin{Verbatim}
    @PolyDet("up") E get(@PolyDet List<E> this, @PolyDet int index);
\end{Verbatim}

@@

\subsection{Type inference, dataflow analysis, and type refinement}\label{dataflow}

\todo{Move all of this to Section 3 since the core calculus has no type
  inference?
  But we do want to discuss \<sort>.  So maybe just move the discussion of type inference?}

The core calculus requires a qualifier on every type.}

\TheDeterminismChecker performs local type inference within method bodies.
It does not perform whole-program type inference.
This makes separate compilation possible.
It forces programmers to write specifications (type qualifiers) on methods,
which is good style and valuable documentation.

@@

In addition to assignments, some method calls refine the types of their
arguments.  One example is the \<sort> routine, as shown in 
as shown in \cref{fig-sorting}.  One notable fact about the \rulename{sort}
rule is that it requires the type argument to be deterministic (type
qualifier \<Det>).  This is necessary because of two constraints.
First, refinement must not change the type argument, because of invariant
type argument subtyping.
Second, the only possible element type for a \<Det> collection is \<Det>.

@@

  $\infer[\rulename{sort}]
  {\Gamma \vdash x_{\mathrm{post}} : \<Det List>\angles{\<Det>\ \beta}}
  {\Gamma \vdash x_{\mathrm{pre}} : \<OrderNonDet List>\angles{\<Det>\ \beta} & \<sort(>x_{\mathrm{pre}}\<)>}$

@@

    \caption{Sorting type refinement rules.  \<sort> is a method that works
      by side effect and changes the value---and the type---of its receiver.
    ``$x_{\mathrm{pre}}$ stands for a value before the call to \<sort>, and
      ``$x_{\mathrm{post}}$ stands for the value after the call.
    }
    \label{fig-sorting}
\end{figure}
As an example, if \codeid{sort()} is called on a receiver of type
\codeid{OrderNonDet List<Det Set<Det String>>}, it will be type refined to
\codeid{Det List<Det Set<Det String>>}. However, it is not permitted to
call \<sort> on a receiver of type \codeid{OrderNonDet List<OrderNonDet
  Set<Det String>>}.  The rule is not applicable, because doing so would
instantiate invalid types.

As another example, it is only sometimes permitted to call \<shuffle> on a receiver of
type \ptype{Det List}{Det String}.  It is illegal to call \<shuffle> on a
receiver of \emph{declared} type \ptype{Det List}{Det String}, because doing so would change
the receiver's type to be inconsistent with its declaration.  However, it
is permitted to call \<shuffle> on a receiver of type \ptype{Det List}{Det
  String} that was declared as \ptype{OrderNonDet List}{Det
  String} but then refined to \ptype{Det List}{Det String}, for example by
calling \<sort> or via assignment.
